---
title: "Inequalities"
order: 1
---

## Reducing Inequalities in Activity Rates

### Background

Health inequalities are unfair, avoidable group-level differences in health outcomes.\
All NHS providers have a duty to work to reduce these inequalities in future, and the [New Hospitals Programme](https://www.gov.uk/government/publications/new-hospital-programme-review-outcome/new-hospital-programme-equality-impact-assessment#:~:text=Women%2C%20children%2C%20the%20elderly%2C,use%20hospital%20services%20in%20general) allows schemes to model the effect of addressing inequalities, which will prevent them from building a hospital which is too small once health inequalities have been addressed.

When predicting future demand for acute care, we have the choice of whether to assume existing inequalities in acute care activity will persist, or to make assumptions around their reduction. If we assume that inequality in activity rates will be rectified (an upward slope of activity rates against deprivation being flattened to become horizontal) by the horizon year, we can make further assumptions about how this change may come about. This is agnostic to the specific ways in which schemes will choose to modify their activity to achieve this aim, but it does assume that any changes made will be successful in removing group-specific inequality. This represents an assumption of the successful application of [proportional universalism](https://www.instituteofhealthequity.org/resources-reports/towards-health-equity-a-framework-for-the-application-of-proportionate-universalism).

### Method

Within each Healthcare Resource Group (HRG) in each scheme, the activity rate in the baseline year is plotted against deciles of [Index of Multiple Deprivation](https://www.gov.uk/government/collections/english-indices-of-deprivation) (IMD[^1] , based on the [Lower Super Output Area](https://www.ons.gov.uk/methodology/geography/ukgeographies/statisticalgeographies) (LSOA) of patient residence, where quintile 1 is the most deprived 10% of LSOAs) and a straight line is fitted[^2] through the ten points. This rate is calculated as the count of that activity per decile of deprivation in the baseline year, divided by the estimated catchment population for the hospital that resides in that deprivation decile. If the line trends upwards from decile 1 to 10, this HRG is added to the list of those where significant inequality was evident[^3]. See @fig-1 for a visual depiction.

[^1]: [IMD19](https://www.gov.uk/government/statistics/english-indices-of-deprivation-2019) in the current model version.

[^2]: Method uses weighted least squares for heteroskedastic errors, where weighting is by [hospital catchment population size](https://csucloudservices-my.sharepoint.com/personal/claire_welsh_mlcsu_nhs_uk/Documents/Documents/NHP/sprint%20info/link%20to%20our%20method%20on%20this) by IMD decile.

[^3]: HRG added to the selectable list only if: - at least 3 IMD quintiles are represented in the baseline data, - counts of HRG activity must exceed 10, and - the statistical significance (p-value) of the slope parameter of the fitted line is \<0.05 with a linear gradient \>0.

```{r}
#| echo: false
#| fig-alt: |
#|  A diagram showing the an example of activity rate plotted against deprivation decile
#| label: fig-1
#| warning: false


library(ggplot2)

#set up some dummy data
data <- data.frame(activity = c(0.0008, 0.0011, 0.0015, 0.0019,0.0020, 0.0024, 0.0020, 0.0031, 0.0033, 0.0035),
                   decile = seq(1,10,1))

#create a linear model to plot and predict from
model <- lm(activity ~ decile, data = data)
upper <- predict(model, newdata = data.frame(decile = 10))
lower <- predict(model, newdata = data.frame(decile = 1))
mid <- mean(data$activity)

#plotting
plot <- ggplot(data, aes(x = as.factor(decile), y = activity)) +
  geom_point(size = 3, colour = "blue") +
  geom_smooth(aes(x = as.numeric(decile), y = activity), method = "lm", 
              colour = "orange", linetype = "dashed", se = F) +
  theme(panel.grid.major = element_blank(),
        panel.grid.minor = element_blank(),
        axis.text = element_text(size = 14),
        text = element_text(size = 14)) +
  theme_bw() +
  labs(y = "Activity rate",
       x = "Deprivation Decile",
       title = "Figure 1. An example of deprivation-related inequality in activity rate",
       subtitle = "Blue dots represent average activity rates per decile.\nThe orange dashed line is a line of best fit through these points.") +
  scale_y_continuous(breaks = seq(0.000,0.004, 0.0005))

suppressMessages(print(plot))

```

For each HRG where significant inequality in activity rates in the baseline year is found, the user has the option to decide on the probability that removal of this inequality will occur by the horizon year, via each of the following options (and submits the percentage chance of each):

A.  **‘Level-up’**: by the horizon year, the activity rate for this HRG will have increased in all deciles to match the fitted activity rate in decile 10 (the least deprived group).
B.  **‘Level-down’**: by the horizon year, the activity rate for this HRG will have reduced in all deciles to the fitted rate of the most deprived decile (decile 1, with a minimum activity rate of zero).
C.  **‘Zero-sum redistribution’**: by the horizon year, the activity rate for this HRG in all deciles will have increased or decreased to match the median activity rate across the whole HRG. This means that activity rates across all deciles will be the same, and that total levels of activity remain the same as in the baseline year.
D.  **No change**: Inequality in delivery for this HRG will remain as it is the baseline year.

For each run of the Monte-Carlo simulation, one of these options will be chosen based on the percentages assigned to them by the user for that HRG. In effect, with these options, the model will allow users to say, “For this HRG, by the horizon year we predict that inequality of activity in this HRG across deprivation deciles will be removed. The likelihood of achieving this via each method is as such”.

@fig-2 below depicts these options.

```{r}
#| fig-alt: |
#|   Visual description of three assumptions available when inequality reduction is predicted for an HRG. 
#|   Each is assigned a percentage to indicate the likelihood of inequalities in 
#|   the HRG being reduced via that option.
#| echo: false
#| warning: false
#| label: fig-2
#| fig-height: 9

library(patchwork)

A <- plot + 
  labs(subtitle = NULL)+
  geom_hline(yintercept = upper, colour = "darkgreen", linewidth = 1.5) +
  geom_segment(data = subset(data, decile !=10),
    aes(x = as.factor(decile), y = activity, xend = as.factor(decile),
                   yend = upper),
               arrow = arrow()) +
  labs(title = "A. Level-up") 


B <- plot + 
  labs(subtitle = NULL)+
  geom_hline(yintercept = lower, colour = "darkgreen", linewidth = 1.5) +
  geom_segment(data = subset(data, decile !=1),
               aes(x = as.factor(decile), 
                   y = activity, 
                   xend = as.factor(decile),
                   yend = lower),
               arrow = arrow()) +
  labs(title = "B. Level-down") 

C <- plot + 
  labs(subtitle = NULL)+
  geom_hline(yintercept = mid, colour = "darkgreen", linewidth = 1.5) +
  geom_segment(aes(x = as.factor(decile), 
                   y = activity, 
                   xend = as.factor(decile),
                   yend = mid),
               arrow = arrow()) +
  labs(title = "C. Zero-sum redistribution") 

fig2 <- A/B/C + plot_layout(heights = unit(5, "cm"))
suppressMessages(print(fig2))

```
