---
title: "Planning Roadmap"
order: 1
---

The roadmap is for anyone who needs to know about planned changes to the 
demand model. It contains high-level indicative areas of development that may
be subject to change in timing and/or scope as programme priorities shift.

This roadmap is updated quarterly.

### Last updated: June 2026

## Recently Completed 
(for full details see [model updates](https://connect.strategyunitwm.nhs.uk/nhp/project_information/project_plan_and_summary/model_updates.html))


```{r}
#| results: asis
#| echo: false
#| warning: false
#| message: false
# Read pin
board <- pins::board_connect()
data <- pins::pin_read(board, name = "claire.welsh/roadmap_full_list")

library(dplyr)
done <- data |>
  filter(Safe.to.show.to.public. == "Yes" &
         (Workstream_1 == "NHP"|Workstream_2 == "NHP") &
         Include.in.NHP.public.roadmap. == 1 &
         NNL == "Done") |>
  select(Public_wording) 
cat(paste0("- ", done$Public_wording, collapse = "\n"))

```

## Working on Now

```{r}
#| results: asis
#| echo: false
#| warning: false
#| message: false

now <- data |>
  filter(Safe.to.show.to.public. == "Yes",
         (Workstream_1 == "NHP"|Workstream_2 == "NHP"),
         Include.in.NHP.public.roadmap. == 1,
         NNL == "Now") |>
  select(Public_wording) 
cat(paste0("- ", now$Public_wording, collapse = "\n"))
```

## Working on Next

```{r}
#| results: asis
#| echo: false
#| warning: false
#| message: false
nnl_next <- data |>
  filter(Safe.to.show.to.public. == "Yes",
         (Workstream_1 == "NHP"|Workstream_2 == "NHP"),
         Include.in.NHP.public.roadmap. == 1,
         NNL == "Next") |>
  select(Public_wording) 
cat(paste0("- ", nnl_next$Public_wording, collapse = "\n"))
```

## Exploring

```{r}
#| results: asis
#| echo: false
#| warning: false
#| message: false
exploring <- data |>
  filter(Safe.to.show.to.public. == "Yes",
         (Workstream_1 == "NHP"|Workstream_2 == "NHP"),
         Include.in.NHP.public.roadmap. == 1,
         NNL == "Later") |>
  select(Public_wording) 
cat(paste0("- ", exploring$Public_wording, collapse = "\n"))
```
